Structural Breaks in Inflation Dynamics within the European Monetary Union

### 2011-04-27 pkg/paper/Paper/draft.Rnw
updated paper and papers.bib results4 treats with seasonal effects
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/glogis@80 b679b2f7-6539-45d0-a605-5ddcb976cde9

<<preliminaries>>=
## packages
if(Sys.getlocale() != "C") Sys.setlocale("LC_TIME", "English")
library("glogis")
library("fxregime")
library("lattice")

## additional functions
moments <- function(object, ...) UseMethod("moments")
moments.glogisfit <- function(object, ...) object$moments
moments.breakpoints.glogisfit <- function(object, breaks = NULL, ...)
  t(sapply(refit(object, breaks = breaks), "[[", "moments"))

## For the creation of the tables
## <<tabelle, eval=FALSE>>=
root_dir <- do.call("file.path", as.list(head(strsplit(getwd(), .Platform$file.sep)[[1]], -1)))
if(!file.exists(file.path(root_dir, "Analysis", "Austria.rda"))) {
setwd(file.path(root_dir, "Analysis"))
## run results3 (with dummy creation for sales periods)
Sweave("results3.Rnw")
setwd(file.path(root_dir, "Paper"))
}

@@

@@

get_data <- function(i) {
  rval <- do.call("merge",
    lapply(i, function(j) results[[j]]$data))
  colnames(rval) <- names(results)[i]
  rval
}

@@

%(... Jesus: journals, EE (empirical economics): mehr EMU, weniger Methode, Journal of Applied Econometrics: genau umgekehrt...)
%(... economic story, big transition ...) \\
Ever since the beginning of the European Union, the topic of a common currency was a controversial issue. Although the Economic and Monetary Union (EMU) is now a fact, the discussion about the economic effects of the Euro is far from being settled. The controversial topics range from the question of whether or not the Euro--zone is indeed an optimal currency area (OCA) all the way to the very survival of the Euro in light of the budgetary problems of a number of its member states. The effects of monetary unions on a number of macroeconomic indicators, with inflation being the most important one in this context is in the center of an ongoing debate, for instance the question about short-run and steady-state inflation uncertainty dealt with in \citet{cap}, or the structural convergence of the inflation rates in EU countries which is the topic of an investigation by \citet{palomba}. \\
%(... big transition continued, our contribution...) \\
The way towards the EMU basically consisted of 3 stages: stage I (1990-1994) as a phase of liberalization, stage II (1995-1998) a phase of convergence and stage III (1999-2001) the transition period, which ended with the introduction of the euro as legal tender. Any new countries who want to join the EMU are first obliged to fulfill the Maastricht which, besides setting rules for government debts and interest rates, also requires the participation in the ERM~II for two years, while the exchange rates towards the euro are not allowed to cross the nominal band. These criteria acknowledge, that a stable inflation rate is one of the key goals of every good macroeconomic policy. \cite{emerson} makes it clear, that a high inflation rate is also more variable and uncertain and thus causes more relative price variability, leading to a less efficient price mechanism.  Failing to stabilize inflation leads to severe economic problems. Throughout the literature, there is still a considerable degree of uncertainty as to what extent the introduction of the euro, or monetary unions in general, affect the inflation rate. There are a number of reasons for suspecting changes. Given that a country experienced quite volatile inflation rates, its efforts to meet the convergence criteria were likely to lead to an alteration at least in the mean (since this was required for a number of countries) and possibly in the volatility of their respective inflation rates as well. The EMU requires a country to possibly change a number of economic policies, which can only be achieved by big transitionally efforts. \\
%(...introduction needs to include the value added of this paper, why is it better, how does it compare to others...) \\
We contribute to this debate by developing a new method for testing for structural breaks in the dynamics of the inflation rates, which focuses on these changes and thereby upon the effect of the EMU on a number of European countries both within and without the Euro--zone. The framework developed in this paper allows testing for any significant changes of a countries inflation dynamics and thus helps interpreting the effects of the EMU. For this purpose, we developed a new estimation approach building on previous research by \citet{z07}. \\
%(...effects of monetary unions, literature overview (more literature in discussion of results)...) \\
Two of the three most important channels that change inflation, the money supply and the degree of openness  are subject to fundamental changes once a country joins a monetary union. The original countries participating in the Exchange Rate Mechanism (ERM) had to synchronize money supply growth and interest rates in order to fix exchanges rates or otherwise risk the chance of severe disequilibria. Any new country willing to join must first participate in the second exchange rate mechanism (ERM~II) with their currencies floating in a rather narrow band against the euro, which basically requires the same synchronization tools. \\
Openness, i.e. the inflow of new goods and services or the abolishment of tariffs and capital restraint and regulations, is more likely to generate deflationary pressure, due to the substitution of cheaper goods for more expensive ones. Since deflation is rather unusual, with general price increases being the norm, the logical conclusion is that the most important inflationary channel are changes in money supply.
%\footnote{Indeed, a general rise in prices is impossible without a simultaneous increase in money supply.}. 
These changes in turn are influenced by a countries monetary policy and the credit behavior of its banking industry. External shocks, like the spillover effects of a hike in oil prices or the financial crisis may be important as well. \\
Economic theory unfortunately does not provide any clear foundation for a proper answer whether or not the creation of a monetary union between two or more states is likely to reduce or increase the variability or even the level of the inflation rate.  
%\citet{cooper} points out, that ``a central bank under a monetary union will internalize the interdependence between countries and optimally choose a lower inflation rate" and he argues that a ``central Bank governing the growth of money supply will optimally choose zero inflation." This is not the case with the ECB which targets at 2\% so as to avoid the alleged risk of deflation. It is thus not quite clear how a monetary union will affect the volatility and the level of the inflation rate.\\
An interesting approach to this question is taken by \citet{holte}. 
%He creates a two country model for monetary policy analysis along the line of two models by \citet{mc1} and \citet{mc2} as well as \citet{gali}. The inflation is modeled by a hybrid Phillips curve  specification and there are a number of different home country interest rate rules, like strict inflation targeting, flexible targeting, pegging to a currency and --  monetary union. 
What he finds out via simulations of  different interest rate rules is that the standard deviation of the home CPI inflation rate can be substantially reduced by joining a monetary union. The effects of joining a monetary union on inflation variability in his framework depend on structural parameters like risk aversion, price flexibility, export demand elasticity, openness and shock correlations. However, due to the fact that not all of these parameters are known and that their interaction as well has to be estimated, the whole model is very dependent on a variety of assumptions. \\
The remainder of this paper is structured as follows: section \ref{s2} presents the data, section \ref{s3} presents the model and the estimation techniques used, section \ref{s4} illustrates our approach using Slovenia as an example, section \ref{s5} presents the results and section \ref{s6} concludes.

@@

%(... Jesus would present data later, first describe what we are going to to do...) \\
HICP series\footnote{seasonally unadjusted} from Jan.1990--Mar.2010 are obtained for 21 countries the \citet{oecd}. Countries included are Austria,	Belgium, Czech Republic,	Denmark, Estonia,	Finland,	France,	Germany,	Greece,	Hungary,	Ireland,	Italy,	Luxembourg,	Netherlands,	Poland,	Portugal,	Slovakia, Slovenia,  Spain,	Sweden, and	United Kingdom. Latvia and Lithuania as well as Bulgaria and Romania are excluded due to data scarcity. Cypria and Malta are not included since they are very small economies. \\
The countries in this sample can be divided into three different groups. First the euro countries (Austria, Belgium, Estonia, France, Finland, Germany, Greece, Ireland, Italy, Luxembourg, the Netherlands, Portugal, Slovakia, Slovenia and Spain).  Second, the EU members not participating in the ERM~II: Czech Republic, Hungary, Poland, United Kingdom, and Sweden. Denmark stands on its own as a member of the EU and the ERM~II, but not yet a member of the EMU. \\

% greece: -for winter discounts: First 10 days of January-end of February; -for summer discounts: First 10 days of July -end od August; but has always been included

@@

% generell strategie: estimate, test parameter stability, breaks model

%neuer Titel
%Strukturbruch
%slowenisches Beispiel

% laut achim noch zuviel bla bla im text --> weniger als anektode, präziser
%(...less anectodic, more precise...) \\
To monitor changes in the dynamics of the inflation rates, we follow the usual strategy employed when testing for structural breaks. First, the model is fitted for the whole sample using the generalized logistic distribution. This then enables testing for the stability of the parameter estimates. If there is evidence of significant changes, one proceeds by computing the number and the timing of the break points. \\
This framework allows to trace the evolution of mean, variance and skewness of the inflation series and thus interpreting deviations in any of these parameters as possibly due to a change in monetary policy, is made feasible. Any huge intervention or change is captured by fluctuations in the parameter stability of the series and can be quantified and related to a distinct time frame. \\
%(...discuss correlation...) \\
The correlation dynamics of the series is of less importance to this framework, which concentrates on changes in mean, variance and skewness and does not focus on autocorrelation. 
%To provide consistent estimates of the parameter vector $\hat{\phi}$, the covariance matrix $\hat{V}^{-1/2}$ is estimated using a heteroskedasticity and autocorrelation consistent covariance estimator.

@@

%(... formulation problems ...) \\
%(...Jesus: GL not standard, explain why it is worth using it, what makes it better than ARIMA?...) \\
%Motivation Verteilung
The data suggested that the use of a normal distribution might cause some estimation problems, since most of the time series exhibit asymmetric properties (with sometimes rather strong skewness) and -- compared to the normal distribution -- a fatter tail. To accommodate for this fact, the use of the generalized logistic distribution (GL) with its relative low degree of complexity, seemed appropriate. On the one hand, it is a simple distribution and on the other hand, it is flexible enough to capture essential properties of the data. For the inflation series $y_t = 100\cdot log(HICP_t/HICP_{t-1})$ we therefore assume a GL distribution as defined in \citet{johnson}:

@@

with location $\theta$, scale $\sigma$ and shape $\delta$. For $\delta=1$ the distribution simplifies to the logistic distribution, for $\delta<1$ it is skewed to the left and for $\delta$$>$1 it is skewed to the right (see appendix \ref{glscore} for more details). \\
In econometrics, the logistic distribution is often used in income distributions and growth models. This is due to  its fatter tail, which fit these kind of data somewhat better. \citet{won} employ a GL distribution in a regression model with autocorrelated errors. They use this type of distribution rather than a Student's t-distribution, as to model the fact that these are oftentimes  non symmetric and severely left or right skewed. A similar GL distribution is also used in \citet{tolikas} who analyze extreme risk and value--at--risk in the German stock market, although they do not use a shape parameter. Regarding inflation rates, the GL distribution is --- to our best knowledge --- only used in context to expected inflation. \citet{batchelor} use a logistic distribution (not its generalization) to model the distribution of mean expected inflation rates. \\

@@

% anders erzählen: um ein Modell zu anzupassen, dann den restlichen teil
%Schätzung
%obiger teil erst später; den verweis auf R erst am Ende in den computational details; auch das zeug ab an enhancement... in die computatinal details aufnehmen
%(... better separation between estimation and test, rest to computational details...) \\

Under the assumption, that $y_t$ is distributed according to the generalized logistic distribution, we test for the hypothesis that the parameter vector $\hat{\phi}$ stays stable over time:

@@

where $\hat{V}^{-1/2}$ is some consistent estimator of the variance of the scores, which is used to decorrelate the series as to avoid problems due to serial correlation. Then a functional central limit theorem (FCLT) for $\mathit{efp}(\cdot)$ is employed, which converges to a 3-dimensional Brownian bridge:

@@

\subsection{Test}
%(.. Jesus would cite a paper by Andrews and Ploberger 1993 that provided the key for creating confidence intervals around the estimates...) \\

After the empirical fluctuation process is computed, it conveys information about departures from the $H_0$ of parameter stability over time. For a precise quantification of the actual deviation from its hypothesized zero--mean path, the main idea of the test statistics presented here is to aggregate the $\mathit{efp}(\cdot)$ in such a way as to be able to construct a test statistics so that critical values and p values can be derived. In this paper, we employ the Supremum of LM (supLM) test by \citet{andrews}. This test is well suited for single break alternatives and performs better if more than just one of the distribution parameters change. It is given by:

@@

\subsection{Breakpoint Estimation}

If a break is detected via the supLM test, the next step is estimating the number and the timing of these break points. Breakpoints $\tau_1,...,\tau_B$ are estimated via maximization of the full segmented likelihood:

@@

% noch mehr dazu: finds 0 wenn nicht signifikant, 1 wenn signifikant, dann auswahl über das lwz criterion

\clearpage
\section{An Example}\label{s4}

% anders: the data encompasses, thus not included....
%(... shortend, changed formulation...) \\
For an illustration of the method, consider the case of Slovenia. The data collected encompasses the period from Jan 1996 onwards. During the beginning of this period, Slovenia experienced very high inflation rates but was successfull in continually decreasing them. A glance at the data suggests a break around 2003/04, with an evident change in the mean.

\begin{figure}[h!tp]
<<fig=TRUE, height=4, width=8>>=
x <- results$Slovenia
par(mfrow = c(1, 2))
plot(x$data, main = "Slovenia", ylab = "100 * diff(log(HICP))", xlab = "Time")
lines(x$breakpoints, breaks = x$nbreaks)
lines(fitted(x$breakpoints, breaks = x$nbreaks, type = "mean"), col = 4)
plot(x$efp, functional = supLM(0.1), main = "supLM test")
lines(x$breakpoints, breaks = x$nbreakss)
@
\caption{Slovenia: breakpoint estimate and supLM test}
\end{figure}

The supLM test confirms this belief by picturing a hike of the $efp(\cdot)$ around 2003. The LWZ criterion suggests one break point, which is estimated at August 2003. \\
In this year, Slovenia introduced a number of financial reforms and severely decreased the year on year growth of money supply. Combined, these actions were enough to ensure the succesfull participation in the ERM~II a year later.

%\section{Power Analysis}
%
%(...Jesus would not do this ...) \\
%
%Basis idea: show that with a GL instead of a Normal, the probability of detecting minimal effects in the series (which is GL--distributed) is higher. If for example, just the skewness parameter changes, is a test build on the normality assumption able to capture this? (...more to come if we decide to include this. In principle, we could then argue in section \ref{s31} that a power analysis suggested that we might miss changes in the third moment or so, i.e. normal has lower power if series is GL--distributed)...)

@@

% besser formulieren

%
%das package colortbl (\usepackege[dvips]{colortbl}) ermöglicht z.B. mit
%\rowcolor[gray]{0.90} in einer Tabellenzeile diese farbig (bzw. hier grau) zu unterlegen.

\footnotesize
\begin{longtable}{|l|l|r|r|r|l|l|l|}
\hline
\multicolumn{8}{|c|}{The Eastern Countries} \\
\hline
Country & Segment & Mean & Variance & Skewness & ERM & ERM~II & Euro \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(3,5,10,15,17,18), format_moments)))
@
\hline
\multicolumn{8}{|c|}{Italy and Spain} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(12, 19), format_moments)))
@
\hline
\multicolumn{8}{|c|}{Ireland} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(11), format_moments)))
@
\hline
\multicolumn{8}{|c|}{No change countries} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(9,14,7,6), format_moments)))
@
\hline
\multicolumn{8}{|c|}{Non participating countries} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(20, 21), format_moments)))
@
\hline
\multicolumn{8}{|c|}{The other countries} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(1,2,4,8,13,16), format_moments)))
@
\caption{Break point and parameter estimates and ERM information}
\end{longtable}
% results3 hier einbauen
% warning: the groups then change; Italy and Greece have not yet answered
% Italian Statistics: Your request has been correctly inserted in the system with the reference number 13078/2010.
% comment: the greek break is NOT sure, since they have not yet answered about sales periods

% resultate: besser ordnen; mit paragraph oder subsection*{}; text über sales do inne; indicate implications in mean and discuss changes in var and skewness

>>>>

<<preliminaries>>=
## packages
if(Sys.getlocale() != "C") Sys.setlocale("LC_TIME", "English")
library("glogis")
library("fxregime")
library("lattice")
library("moments")

## additional functions
momentgen <- function(flag=1,x, start=yearmon(c(1990,1998,2007)), end=yearmon(c(1998+11/12,2007+11/12,2010+11/12))) {
# 1:mean, 2:variance, 3:skewness
if(flag==1){
mean = mat.or.vec(ncol(x),length(start))
for(jj in 1:length(start)){
x1 <- window(x, start=start[jj], end=end[jj])
for(i in 1:ncol(x1)) {
mean[i,jj] = mean(na.omit(x1[,i]))
}
}
ret1 <- c(quantile(mean[,1], probs=.25), quantile(mean[,1], probs=.50), quantile(mean[,1], probs=.75))
ret1 <- cbind(ret1)
ret2 <- c(quantile(mean[,2], probs=.25), quantile(mean[,2], probs=.50), quantile(mean[,2], probs=.75))
ret2 <- cbind(ret2)
ret3 <- c(quantile(mean[,3], probs=.25), quantile(mean[,3], probs=.50), quantile(mean[,3], probs=.75))
ret3 <- cbind(ret3)
ret <- cbind(ret1,ret2,ret3)
colnames(ret) <- c(start[1], start[2], start[3])
rownames(ret) <- c("lower", "medium", "upper")
return(ret)
}
if(flag==2){
var = mat.or.vec(ncol(x),length(start))
for(jj in 1:length(start)){
x1 <- window(x, start=start[jj], end=end[jj])
for(i in 1:ncol(x1)) {
var[i,jj] = var(na.omit(x1[,i]))
}
}
ret1 <- c(quantile(var[,1], probs=.25), quantile(var[,1], probs=.50), quantile(var[,1], probs=.75))
ret1 <- cbind(ret1)
ret2 <- c(quantile(var[,2], probs=.25), quantile(var[,2], probs=.50), quantile(var[,2], probs=.75))
ret2 <- cbind(ret2)
ret3 <- c(quantile(var[,3], probs=.25), quantile(var[,3], probs=.50), quantile(var[,3], probs=.75))
ret3 <- cbind(ret3)
ret <- cbind(ret1,ret2,ret3)
colnames(ret) <- c(start[1], start[2], start[3])
rownames(ret) <- c("lower", "medium", "upper")
return(ret)
}
if(flag==3){
skewness= mat.or.vec(ncol(x),length(start))
for(jj in 1:length(start)){
x1 <- window(x, start=start[jj], end=end[jj])
for(i in 1:ncol(x1)) {
skewness[i,jj] = skewness(na.omit(x1[,i]))
}
}
ret1 <- c(quantile(skewness[,1], probs=.25), quantile(skewness[,1], probs=.50), quantile(skewness[,1], probs=.75))
ret1 <- cbind(ret1)
ret2 <- c(quantile(skewness[,2], probs=.25), quantile(skewness[,2], probs=.50), quantile(skewness[,2], probs=.75))
ret2 <- cbind(ret2)
ret3 <- c(quantile(skewness[,3], probs=.25), quantile(skewness[,3], probs=.50), quantile(skewness[,3], probs=.75))
ret3 <- cbind(ret3)
ret <- cbind(ret1,ret2,ret3)
colnames(ret) <- c(start[1], start[2], start[3])
rownames(ret) <- c("lower", "medium", "upper")
return(ret)
}
}

moments <- function(object, ...) UseMethod("moments")
moments.glogisfit <- function(object, ...) object$moments
moments.breakpoints.glogisfit <- function(object, breaks = NULL, ...)
  t(sapply(refit(object, breaks = breaks), "[[", "moments"))

## For the creation of the tables
## <<tabelle, eval=FALSE>>=
root_dir <- do.call("file.path", as.list(head(strsplit(getwd(), .Platform$file.sep)[[1]], -1)))
if(!file.exists(file.path(root_dir, "Analysis", "Austria.rda"))) {
setwd(file.path(root_dir, "Analysis"))
## run results4 (with dummy creation for sales periods)
Sweave("results4.Rnw")
setwd(file.path(root_dir, "Paper"))
}

@@

table2<- vector(length = length(fnam), mode = "list")
names(table2) <- sapply(strsplit(sapply(strsplit(fnam, "/", fixed = TRUE), tail, 1), ".", fixed = TRUE), head, 1)

for(qq in seq_along(fnam)) {
  load(file.path(root_dir, "Analysis", fnam[qq]))
  table2[[qq]] <- x
}

@@

format_momenteuro <- function(momeuro, digits = 6, name="Euro") {
  rval <- format(round(momeuro, digits = digits), nsmall = digits)
  rval <- cbind(name,"", matrix(paste("$", rval, "$", sep = ""), ncol = 3))
  rval[1,1] <- name
  rval[2,1] <- ""
  rval[3,1] <- ""
  rval[1,2] <- "lower"
  rval[2,2] <- "medium"
  rval[3,2] <- "upper"
  rval <- apply(rval, 1, paste, collapse = " & ")
  rval <- paste(rval, "\\\\")
  rval[length(rval)] <- paste(rval[length(rval)], "\\hline")
  return(rval)
}
get_data <- function(i) {
  rval <- do.call("merge",
    lapply(i, function(j) results[[j]]$data))
  colnames(rval) <- names(results)[i]
  rval
}

@@

Ever since the beginning of the European Union, the topic of a common currency was a controversial issue. Although the Economic and Monetary Union (EMU) is now a fact, the discussion about the economic effects of the Euro is far from being settled. The controversial topics range from the question of whether or not the Eurozone is indeed an optimal currency area (OCA) all the way to the very survival of the Euro in light of the budgetary problems of some of its member states. The effects of monetary unions on a number of macroeconomic indicators, with inflation being the most important one in this context, is in the center of an ongoing debate: this regards the question about short-run and steady-state inflation uncertainty dealt with in \citet{cap}, or the structural convergence of the inflation rates in EU countries, which is the topic of an investigation by \citet{palomba}. \\
The way towards the EMU basically consisted of 3 stages: stage I (1990-1994) as a phase of liberalization, stage II (1995-1998) a phase of convergence and stage III (1999-2001) the transition period, which ended with the introduction of the euro as legal tender. A possible fourth stage is the continuing integration of new member states, mostly former communist countries. Any new state that wants to join the EMU is first obliged to fulfill the Maastricht criteria, which, besides setting rules for government debts and interest rates, also requires the participation in the ERM~II (Exchange Rate Mechanism~II) for two years, while the exchange rates towards the euro are not allowed to cross the nominal band.\\
\cite{emerson} put emphasize on the fact that a high inflation rate is also more variable and uncertain and thus causes more relative price variability, leading to a less efficient price mechanism.  Failing to stabilize inflation leads to severe economic problems. Throughout the literature, there is still a considerable degree of uncertainty as to what extent the introduction of the euro, or monetary unions in general, affect the inflation rate. The EMU requires a country to possibly alter a number of economic policies and institutions, which may demand for significant short term adjustments to take place. 
Given that a country experienced quite volatile inflation rates, its efforts to meet the convergence criteria are likely to lead to an alteration at least in the mean (since this was one of the requirements) and possibly in the volatility of their respective inflation rates as well. \\
%(...too short: need: methodological ideas and main questions..., value added: schiefe, schwere ränder)\\
% http://home.iitk.ac.in/~kundu/paper152.pdf  (gl-beschreibung in unterlagen), p.4 more peaked in center, but heavier tails
However, a deeper inquiry into this problem is in need of a method able to capture changes in the inflation dynamics. We contribute to this debate by developing a new method for testing for structural breaks in the dynamics of the inflation rates, which focuses on these changes and thereby upon the effect of the EMU on a number of European countries both within and without the Eurozone. The framework developed in this paper makes it possible to test for any significant changes of a countries inflation dynamics and thus helps interpreting the effects of the EMU. We focus on changes in the first three moments of these inflation rates, which we assume to be distributed according to a generalized logistic distribution (GL), as defined in Section \ref{s3}. The GL distribution is a suitable one due to its flexibility to capture nonsymmetry and because of its fatter tails. For this purpose, we developed a new estimation approach building on previous research by \citet{z07}. \\
%(... am Schluss so enden, dass nun interessant, wie sich die Momente so genau verändert haben; auf Basis der theorie--> suche nach brüchen, schiefe noch hinein/besser hinein; 3 saha simultan aschoa...)\\
This framework allows us to observe the evolution of mean, variance and skewness over time. 
%Considering the prior knowledge available about the timing of a countries participation in any of the aforementioned steps towards monetary union, an exact timing of breakpoints in the moments of the respective inflation series enables us to draw conclusions about the exact effects of the EMU the inflation rates of its members. 
Our estimation technique enables the simultaneous estimation of breaks in the mean (which is important in the context of the Maastricht criteria), the variance (which should be low for the price mechanism to be efficient) and the skewness (where a change in skewness is evidence of a shift of the distribution towards a new ``center of gravity'') of the inflation rates of interest. This than can be used to search for answers regarding the question about the effect of the EMU on the inflation dynamics of its member states or more general to investigate the effects of external shocks.\\
The remainder of this paper is structured as follows: Section \ref{s2} presents the data, Section \ref{s3} presents the model and the estimation techniques used, Section \ref{s4} illustrates our approach using Slovenia as an example, Section \ref{s5} presents the results and Section \ref{s6} concludes.

@@

We employ seasonally unadjusted HICP series of 21 countries ranging from January 1990 to December 2010 that are obtained from the \citet{oecd}. Countries included are Austria,	Belgium, Czech Republic,	Denmark, Estonia,	Finland,	France,	Germany,	Greece,	Hungary,	Ireland,	Italy,	Luxembourg,	Netherlands,	Poland,	Portugal,	Slovakia, Slovenia,  Spain,	Sweden, and	United Kingdom. Latvia and Lithuania as well as Bulgaria and Romania are excluded due to data scarcity. Cypria and Malta are not included since they are very small economies. \\
The countries in this sample can be divided into three different groups. First the euro countries (Austria, Belgium, Estonia, France, Finland, Germany, Greece, Ireland, Italy, Luxembourg, Netherlands, Portugal, Slovakia, Slovenia and Spain).  Second, the EU members not participating in the ERM~II: Czech Republic, Hungary, Poland, United Kingdom, and Sweden. Denmark stands on its own as a member of the EU and the ERM~II, but not yet a member of the EMU. \\
%(... schlussverkäufe schon hier erwähnen, einfach cool, ja das gibts, wir haben die aber weggeschätzt; generell viel angelsächsischer/präpotenter schreiben; nur 1 weg der richtige und zwar unserer; nicht erwähnen, dass andere dinge auch noch möglich; so nach motto: mich interessierts, ich machs, andere erst am ende erwähnen...)
% greece: -for winter discounts: First 10 days of January-end of February; -for summer discounts: First 10 days of July -end od August; but has always been included
In the case of Belgium and Luxembourg, we dealt with the inclusion of the sales periods effects\footnote{This was demanded by a Commission regulation, details may be found in: \citet{eurlex} and in the Appendix \ref{dataprobs}.} into the HICP series (January 2000 for Belgium and January 1999 for Luxembourg) by estimating dummies for these month\footnote{There are two month with a significant decrease in prices (January and July), followed by significant price increases in the following months} and subtracting them from the original time series.

@@

To monitor changes in the dynamics of the inflation rates, we follow the usual strategy employed when testing for structural breaks. First, the model is fitted for the whole sample using the generalized logistic distribution. This then enables testing for the stability of the parameter estimates. If there is evidence of significant changes, the number and the timing of the break points are computed. \\
This framework allows to trace the evolution of mean, variance and skewness of the inflation series and thus interpreting deviations in any of these parameters as possibly due to a change in monetary policy, is made feasible. Any huge intervention or change is captured by fluctuations in the parameter stability of the series and can be quantified and related to a distinct time frame. \\
%(...discuss correlation...) \\
The correlation dynamics of the series is of less importance to this framework, which concentrates on changes in mean, variance and skewness and does not focus on autocorrelation, the use of alternative testes, as described in \cite{andreou}, is also hindered by the low number of observations, due to the monthly frequencies.\footnote{In some years the Billion Prices Project of the \cite{bpp}, which provides a much higher data frequency, could solve this problem.} 
%To provide consistent estimates of the parameter vector $\hat{\phi}$, the covariance matrix $\hat{V}^{-1/2}$ is estimated using a heteroskedasticity and autocorrelation consistent covariance estimator.

@@

The data suggested that the use of a normal distribution might cause some estimation problems, since most of the time series exhibit asymmetric properties (with sometimes rather strong skewness) and -- compared to the normal distribution -- a fatter tail. To accommodate for this fact, the use of the generalized logistic distribution (GL) with its relative low degree of complexity, seemed appropriate. On the one hand, it is a simple distribution and on the other hand, it is flexible enough to capture essential properties of the data. For the inflation series $y_t = 100\cdot log(HICP_t/HICP_{t-1})$ we therefore assume a GL distribution as defined in \citet{johnson}:

@@

with location $\theta$, scale $\sigma$ and shape $\delta$. For $\delta=1$ the distribution simplifies to the logistic distribution, for $\delta<1$ it is skewed to the left and for $\delta$$>$1 it is skewed to the right (see Appendix \ref{glscore} for more details). \\
In econometrics, the logistic distribution is often used in income distributions and growth models. This is due to  its fatter tails, which fit these kind of data somewhat better. \citet{won} employ a GL distribution in a regression model with autocorrelated errors. They use this type of distribution rather than a Student's t-distribution, as to model the fact that these are oftentimes  non symmetric and severely left or right skewed. A similar GL distribution is also used in \citet{tolikas} who analyze extreme risk and value--at--risk in the German stock market, although they do not use a shape parameter. Regarding inflation rates, the GL distribution is --- to our best knowledge --- only used in context with expected inflation. \\
\citet{batchelor} use a logistic distribution (not its generalization) to model the distribution of mean expected inflation rates. \\

@@

Under the assumption that $y_t$ is distributed according to the generalized logistic distribution, we test for the hypothesis that the parameter vector $\hat{\phi}$ stays stable over time:

@@

where $\hat{V}$ is some consistent estimator of the variance of the scores, which is used to decorrelate the series as to avoid problems due to serial correlation. Then a functional central limit theorem (FCLT) for $\mathit{efp}(\cdot)$ is employed, which converges to a 3-dimensional Brownian bridge:

@@

\subsection{Test}

After the empirical fluctuation process is computed, it conveys information about departures from the $H_0$ of parameter stability over time. For a precise quantification of the actual deviation from its hypothesized zero--mean path, the main idea of the test statistics presented here is to aggregate the $\mathit{efp}(\cdot)$ in such a way as to be able to construct a test statistics so that critical values and p values can be derived. In this paper, we employ the Supremum of LM (supLM) test by \citet{andrews}. This test is well suited for single break alternatives and performs better if more than just one of the distribution parameters change.

\clearpage
It is given by:

@@

\subsection{Breakpoint Estimation}
%(...cite bai and perron 2003 paper here, nachlesen, findet achim super, bei slowenien will er noch konfidenzintervalle dazu...)\\
The dating of structural breaks builds upon the work of \cite{bp03}. If a break is detected via the supLM test, the next step is estimating the number and the timing of these break points. Breakpoints $\tau_1,...,\tau_B$ are estimated via maximization of the full segmented likelihood:

@@

\clearpage
\section{An Example}\label{s4}
%(...more text: 1. motivation, 2.clear methodological steps, 3.interpretation...) \\
<<echo=FALSE>>=
        x <- results$Slovenia
        slovpval <- sctest(x$efp, functional = supLM(0.1))$p.value
@
To illustrate our method, we consider the case of Slovenia. After a period of relative instability following the independence from Yugoslavia, Slovenia was successfull in realigning its economy and introduced a number of reforms leading to a stable growth in the past years. The good economic performance made it possible for Slovenia to enter the ERM~II in June 2004 and later on to introduce the Euro in January 2007. The data collected contains the period from January 1996 onwards. In the first few years, Slovenia experienced very high inflation rates, but was successfull in reducing inflation to a much lower level. \\
A glance at the time series depicted in Figure \ref{fig1} of the monthly inflation rates reveals a decline in the mean around the middle of the observed time period. The $efp(\cdot)$ starts to significantly deviate as early as 2001 and reaches the peak somewhere between 2003 and 2004. The p--value of the supLM test is \Sexpr{round(slovpval, 5)} and clearly indicates at least one break.  The number of breaks is calculated by means of the LWZ criteria, which chooses one break in August 2003. In this year, Slovenia introduced a number of financial reforms and severely decreased the year on year growth in the M3 monetary aggregate. Combined, these actions were enough to ensure the successfull participation in the ERM~II a year later and the introduction of the Euro two years later.

\begin{figure}[h!tp]
<<fig=TRUE, height=4, width=8>>=
x <- results$Slovenia
par(mfrow = c(1, 2))
plot(x$data, main = "Slovenia", ylab = "100 * diff(log(HICP))", xlab = "Time")
lines(x$breakpoints, breaks = x$nbreaks)
lines(fitted(x$breakpoints, breaks = x$nbreaks, type = "mean"), col = 4)
plot(x$efp, functional = supLM(0.1), main = "supLM test")
lines(x$breakpoints, breaks = x$nbreakss)
@
\caption{Slovenia: breakpoint estimate and supLM test}
\label{fig1}
\end{figure}

@@

% das package colortbl (\usepackege[dvips]{colortbl}) ermöglicht z.B. mit
%\rowcolor[gray]{0.90} in einer Tabellenzeile diese farbig (bzw. hier grau) zu unterlegen.

\footnotesize
\begin{longtable}{|l|l|r|r|r|l|l|l|}
\hline
\multicolumn{8}{|c|}{Austria and Belgium} \\
\hline
Country & Segment & Mean & Variance & Skewness & ERM & ERM~II & Euro \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(1, 2), format_moments)))
@
\hline
\multicolumn{8}{|c|}{The Eastern countries} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(3,5,10,15,17,18), format_moments)))
@
\hline
\multicolumn{8}{|c|}{Italy and Spain} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(12, 19), format_moments)))
@
\hline
\multicolumn{8}{|c|}{Ireland} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(11), format_moments)))
@
\hline
\multicolumn{8}{|c|}{No change countries} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(9,14,7,6), format_moments)))
@
\hline
\multicolumn{8}{|c|}{The 90s} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(16,20, 21), format_moments)))
@
\hline
\multicolumn{8}{|c|}{Former DM--zone} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(4,8,13), format_moments)))
@
\caption{Break point and parameter estimates and ERM information}
\label{resulttable}
\end{longtable}
% warning: the groups then change; Italy and Greece have not yet answered
% Italian Statistics: Your request has been correctly inserted in the system with the reference number 13078/2010.
% comment: the greek break is NOT sure, since they have not yet answered about sales periods

\clearpage
\normalsize

%(...3 punkte darstellung noch einbauen...) \\
%(...andere struktur: 1.short introduction, 2. countries in detail (6.1), 3. summary (6.2) ...) \\
Table \ref{resulttable} presents the break dates and the parameter values of the segments as well as the dates of entry into the ERM/ERM~II and the date of the euro introduction. We formed seven groups of countries according to the observed pattern of break points. These seven groups consist of Austria and Belgium, the Eastern European countries, the two big southern European countries, Italy and Spain, Ireland, the countries with no change in the inflation dynamics, the countries with breaks in the early 90s and the member countries of the former DM--zone (excluding Belgium). \\

\subsection{The countries in detail}
